% main.tex — IEEE Conference Paper: Voice & Vision Assistant
% Compile with: pdflatex main.tex ; bibtex main ; pdflatex main.tex ; pdflatex main.tex
\documentclass[conference]{IEEEtran}

% ── Packages ──────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{url}
\usepackage{balance}
\usepackage{xcolor}
\usepackage{array}

% ── Figure path ───────────────────────────────────────────────────────────
\graphicspath{{figures/}}

% ── Title & Authors ───────────────────────────────────────────────────────
% NOTE: Author names and affiliations extracted from sample paper 2.pdf.
%       If these are incorrect, edit below or see Papers/author_choices.txt.
\title{Voice \& Vision Assistant: A Real-Time Multimodal Accessibility System for Blind and Visually Impaired Users}

\author{
    \IEEEauthorblockN{%
        Dharmik Dholu\IEEEauthorrefmark{1},
        Krish Gangadiya\IEEEauthorrefmark{1},
        Samarth Mangukiya\IEEEauthorrefmark{1},
        Vansh Kothia\IEEEauthorrefmark{1}
    }
    \IEEEauthorblockA{%
        \IEEEauthorrefmark{1}Department of Computer Engineering,\\
        Marwadi University, Rajkot, Gujarat, India\\
        \textit{Under the Guidance of: Prof. Jignesh Madrakia}\\
        Email: \{dharmik.dholu, krish.gangadiya, samarth.mangukiya, vansh.kothia\}@marwadiuniversity.ac.in
    }
}

\begin{document}
\maketitle

% ── Abstract ──────────────────────────────────────────────────────────────
\begin{abstract}
Blind and visually impaired individuals face persistent barriers in navigation, reading, and social interaction---barriers that existing assistive tools address only partially and with high latency. This paper presents Voice \& Vision Assistant (VVA), an open-source, real-time multimodal system that fuses twelve perception capabilities: YOLOv8n object detection, MiDaS monocular depth estimation, edge-aware segmentation, three-tier OCR, braille recognition, QR/AR scanning, consented face recognition, GCC-PHAT sound localization, CLIP-based action recognition, LLM-driven visual question answering, FAISS-backed retrieval-augmented memory, and low-latency speech synthesis. Orchestrated by an event-driven pipeline with per-stage timeouts, the system achieves median end-to-end perception latency of 62\,ms on an NVIDIA RTX 4060, delivering spoken micro-navigation cues within a 500\,ms speech-to-speech budget. A five-layer architecture (shared, core, application, infrastructure, apps), enforced by automated import-linting, ensures independent testability across 429+ unit, integration, and performance tests. Privacy-first design---local-only FAISS indexing, opt-in face consent, GDPR Art.~17/20 endpoints---makes VVA suitable for healthcare and education deployments. We present the system architecture, methodology, benchmark plans, and a survey of 50 related works. The full codebase, model checkpoints, and test suite are released as open source.
\end{abstract}

% ── Keywords ──────────────────────────────────────────────────────────────
\begin{IEEEkeywords}
assistive technology, visual impairment, multimodal fusion, real-time object detection, monocular depth estimation, voice interaction
\end{IEEEkeywords}

% ── Body ──────────────────────────────────────────────────────────────────
\input{sections/intro}
\input{sections/background}
\input{sections/related_work}
\input{sections/methods}
\input{sections/experiments}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}
\input{sections/acknowledgements}

% ── References ────────────────────────────────────────────────────────────
\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}
